Fast Sorting on Modern Computers

### 2005-05-31 paper/docs/heap.tex
changed optimization spelling

The only other optimisation used is to unroll the loop while destroying the
heap. This has minimal affect, however.

@@

In the base heapsort, an improvement was added that removed the need for a
sentinel. Since the 4- and 8-heaps were unsorted, this optimisation had to be
removed. Leaving the key being copied in place was still possible, but it needs
to be replaced by a maximal key before trying to sift the next key down the
heap, lest it were chosen to be sifted up. Placing the sentinel is still cheaper
than the bounds check, however, as the bounds check will need to be performed
$NlogN$ times, compared to $N$ times for the sentinel. The cost of the sentinel
write is more expensive but the extra $logN$ instructions saved will mean that
this increase will be an order of magnitude lower than the bounds check.

>>>>

The only other optimization used is to unroll the loop while destroying the
heap. This has minimal affect, however.

@@

In the base heapsort, an improvement was added that removed the need for a
sentinel. Since the 4- and 8-heaps were unsorted, this optimization had to be
removed. Leaving the key being copied in place was still possible, but it needs
to be replaced by a maximal key before trying to sift the next key down the
heap, lest it were chosen to be sifted up. Placing the sentinel is still cheaper
than the bounds check, however, as the bounds check will need to be performed
$NlogN$ times, compared to $N$ times for the sentinel. The cost of the sentinel
write is more expensive but the extra $logN$ instructions saved will mean that
this increase will be an order of magnitude lower than the bounds check.

### 2005-05-31 paper/docs/merge.tex
changed optimization spelling

%\section{Description}
\label{merge}
Merging is a technique which takes two sorted lists, and combines them into a
single sorted list. Starting at the smallest key on both lists, it writes the
smaller key to a third list. It then compares the smallest key on one list with
the next smallest on the other, and writes the smaller to the auxiliary list.
The merging continues until both lists are combined into a single sorted list.

Mergesort works by treating an unsorted array of length $N$ as $N$ sorted arrays of
length one. It then merges the first two arrays into one array of length two, and
likewise for all other pairs. These arrays are then merged with their neighbours
into arrays of length four, and so on, until the arrays are completely sorted.

@@

The number of merges required is $log_2(N)$, and if this number is even then
the final merge puts the array in the correct place. Otherwise, the result will
need to be copied back into the initial array.

@@

For his base algorithm, LaMarca takes Knuth's algorithm N \cite{Knuth98}.
LaMarca describes this algorithm as functioning as described above; lists of
length one are merged into lists of length two, then four, eight, and so on.
However, algorithm N does not work in this fashion. Instead, it works as a
\n{natural} merge. Rather than each list having a fixed length at a certain
point, lists are merged according to the initial ordering. If it happens that
there exists an ascending list on one side, this is exploited by the algorithm.
While this results in good performance, the merge is not perfectly regular, and
it is difficult or impossible to perform several of the optimisations LaMarca
describes on it.

@@

Knuth's algorithm S is a \n{straight} merge, and works as described above. It has
15\% lower instruction count than Algorithm N\footnote{This is observed on
random data, which algorithm N is not designed for. On more regular data, or
semi-sorted data, this will probably not occur.}. Algorithm S, like algorithm N,
is a very difficult algorithm to modify. Knuth describes it in assembly, and
making the optimisations that LaMarca describes on it is more difficult than
rewriting the algorithm more simply. The rewritten algorithm, referred to from
here on as `base mergesort', is a simple higher level version of algorithm S,
and performs only slightly more slowly.

@@

\subsection{Base Mergesort}
\label{base mergesort}
Three optimisations are described by LaMarca for the base mergesort. Firstly, a
bitonic sort, as described in \cite{Sedgewick02}, removes instructions from the
inner loop. Every second array is created in reverse, so that the larger keys  
are next to each other in the middle. The smallest keys are at the outside,
meaning that the lists are iterated through from the outside in. As a result, it
is not possible to go outside the array, and it is not necessary to consider the
lengths of the arrays, apart from when setting the initial array indices. When
the first array is exhausted, it will point to the last key on the other array,
which will be larger than all other keys in that array. When both are
exhausted, the last key of both arrays will be equal. A check is put in, in this
case, and merging ends if both indices are the same. Figure \vref{Visual Sort}
shows a bitonic merge.

@@

Overall, these three optimisations reduced the instruction count by 30\%.

@@

While this step makes sense, whether this is valuable is very much a platform
dependant issue. Since virtual addresses are used, it may be possible that the
addresses used by the cache are not aligned as expected. Even worse, an
operating system with an excellent memory management system may do a similar
step itself. The program's attempt to reduce cache misses could then increase
them. In the event that the \n{Memory Management Unit} in the processor uses
virtual addresses to index its cache, or has a direct mapping between virtual
and physical or bus addresses, then this problem is very unlikely. The
simulations presented in this project address the cache by virtual address, and
so this optimisation is effective in this case.

@@

Despite the improvement in cache misses due to multi-mergesort, it still cannot
account for the increase in instruction count as a result, and it has a higher
cycle count on the Pentium 4 (see Figure \ref{merge cycles}) as a result. The
other sorts generally perform in $O(NlogN)$ time, with double tiled mergesort
being the fastest. Note that the alignment of the mergesort may not work as
desired due to the size of the cache. This was optimised for a 2MB cache, while
the Pentium 4 cache is only 256KB. This means that the two arrays are aligned to
the same cache block. However, since the cache is 8-way associative, this is not
serious, and should not result in a lot of thrashing.

>>>>

%\section{Description}
\label{merge}
Merging is a technique which takes two sorted lists, and combines them into a
single sorted list. Starting at the smallest key on both lists, it writes the
smaller key to a third list. It then compares the next smallest key on that list
with the next smallest on the other, and writes the smaller to the auxiliary
list. The merging continues until both lists are combined into a single sorted
list.

Mergesort works by treating an unsorted array of length $N$ as $N$ sorted arrays of
length one. It then merges the first two arrays into one array of length two, and
likewise for all other pairs. These arrays are then merged with their neighbours
into arrays of length four, then eight, and so on until the arrays are
completely sorted.

@@

The number of merges required is $log_2(N)$, and if this number is even then the
final merge puts the array in the correct place. Otherwise, the sorted array
will need to be copied back into the initial array.

@@

For his base algorithm, LaMarca takes Knuth's algorithm N \cite{Knuth98}.
LaMarca describes this algorithm as functioning as described above; lists of
length one are merged into lists of length two, then four, eight, and so on.
However, algorithm N does not work in this fashion. Instead, it works as a
\n{natural} merge. Rather than each list having a fixed length at a certain
point, lists are merged according to the initial ordering. If it happens that
there exists an ascending list on one side, this is exploited by the algorithm.
While this results in good performance, the merge is not perfectly regular, and
it is difficult or impossible to perform several of the optimizations LaMarca
describes on it.

@@

Knuth's algorithm S is a \n{straight} merge, and works as described above. It has
15\% lower instruction count than Algorithm N\footnote{This is observed on
random data, which algorithm N is not designed for. On more regular data, or
semi-sorted data, this will probably not occur.}. Algorithm S, like algorithm N,
is a very difficult algorithm to modify. Knuth describes it in assembly, and
making the optimizations that LaMarca describes on it is more difficult than
rewriting the algorithm more simply. The rewritten algorithm, referred to from
here on as `base mergesort', is a simple higher level version of algorithm S,
and performs only slightly more slowly.

@@

\subsection{Base Mergesort}
\label{base mergesort}
Three optimizations are described by LaMarca for the base mergesort. Firstly, a
bitonic sort, as described in \cite{Sedgewick02}, removes instructions from the
inner loop. Every second array is created in reverse, so that the larger keys  
are next to each other in the middle. The smallest keys are at the outside,
meaning that the lists are iterated through from the outside in. As a result, it
is not possible to go outside the array, and it is not necessary to consider the
lengths of the arrays, apart from when setting the initial array indices. When
the first array is exhausted, it will point to the last key on the other array,
which will be larger than all other keys in that array. When both are
exhausted, the last key of both arrays will be equal. A check is put in, in this
case, and merging ends if both indices are the same. Figure \vref{Visual Sort}
shows a bitonic merge.

@@

Overall, these three optimizations reduced the instruction count by 30\%.

@@

While this step makes sense, whether this is valuable is very much a platform
dependant issue. Since virtual addresses are used, it may be possible that the
addresses used by the cache are not aligned as expected. Even worse, an
operating system with an excellent memory management system may do a similar
step itself. The program's attempt to reduce cache misses could then increase
them. In the event that the \n{Memory Management Unit} in the processor uses
virtual addresses to index its cache, or has a direct mapping between virtual
and physical or bus addresses, then this problem is very unlikely. The
simulations presented in this project address the cache by virtual address, and
so this optimization is effective in this case.

@@

Despite the improvement in cache misses due to multi-mergesort, it still cannot
account for the increase in instruction count as a result, and it has a higher
cycle count on the Pentium 4 (see Figure \ref{merge cycles}) as a result. The
other sorts generally perform in $O(NlogN)$ time, with double tiled mergesort
being the fastest. Note that the alignment of the mergesort may not work as
desired due to the size of the cache. This was optimized for a 2MB cache, while
the Pentium 4 cache is only 256KB. This means that the two arrays are aligned to
the same cache block. However, since the cache is 8-way associative, this is not
serious, and should not result in a lot of thrashing.

### 2005-05-31 paper/docs/method.tex
changed optimization spelling

As a result, it was necessary to use the original version, and to try to perform
optimisations on it. However, several factors made this difficult or impossible.
LaMarca unrolled the inner loop of mergesort, but it is not clear which is the
inner loop. The two deepest loops were executed very infrequently, and
unrolling them had no effect on the instruction count. Unrolling the other loops
also had no effect, as it was rare for the loop to execute more than five or six
times, and once or twice was more usual.

@@

As a result, a mergesort was written partially based on the outline of algorithm
N. The outer loop was very similar, but the number of merges required were
calculated in advance, and the indices of each merge step.  It was very simple
to turn this into a bitonic array; adding pre-sorting reduced the instruction
count, as did unrolling the loop. The reduction in instruction count from the
initial version, which performed the same number of instructions as the same as
algorithm N, and the optimised version, was about 30\%.

>>>>

As a result, it was necessary to use the original version, and to try to perform
optimizations on it. However, several factors made this difficult or impossible.
LaMarca unrolled the inner loop of mergesort, but it is not clear which is the
inner loop. The two deepest loops were executed very infrequently, and
unrolling them had no effect on the instruction count. Unrolling the other loops
also had no effect, as it was rare for the loop to execute more than five or six
times, and once or twice was more usual.

@@

As a result, a mergesort was written partially based on the outline of algorithm
N. The outer loop was very similar, but the number of merges required were
calculated in advance, and the indices of each merge step.  It was very simple
to turn this into a bitonic array; adding pre-sorting reduced the instruction
count, as did unrolling the loop. The reduction in instruction count from the
initial version, which performed the same number of instructions as the same as
algorithm N, and the optimized version, was about 30\%.
